Public Security Concern: Bicycle Theft in Toronto Shows a Seasonal and Geographical Pattern

### 2023-11-13 outputs/paper/paper.qmd
Change figures and add discussion part

The remainder of this paper is structured as follows: @sec-data discusses the data with @sec-data_introduction includes basic information about data collection and cleaning results, and @sec-data_visualization includes graphs and tables representing relationships between variables and some discussions.

@@

@fig-yearly_count_by_month shows monthly number of offence for each year from 2014 to 2022. The bar plot provides a clear demonstration for not only the overall trend of bicycle thefts through years, but also the potential seasonal pattern that could be further explored.  It shows that the total number of offence in a year increased before 2018 and began decrease since 2018. The change is relatively not significant within 1000, and the number of offence in 2022 is similar to that in 2014. Possible reasons include improvement of security equipment of the bike, the market behind bicycle theft influenced by pandemic, and people's usage habit of bike. These implications need more data to analyze and could be our further interest.

@fig-yearly_count_by_month also illustrates potential seasonal pattern that there are much fewer bicycle thefts in winter, eg. December, January, and February. The offence increases a lot in other seasons especially in summer, eg. June, July, and August. This may be related to the weather in Toronto that many people will not ride bikes in winter due to extremely freezing weather and frequent snowfall.

@@

data |> 
  mutate(occ_month = factor(occ_month, levels = month.name)) |> 
  ggplot(aes(x=as.factor(occ_year), fill=occ_month)) +
  geom_bar() +
  labs(x = "Occurrence Year", y = "Number of occurrence", fill = "Occurrence Month") +
  theme_minimal()
```

Other than seasonal pattern, @fig-location_count_by_year also demonstrates the relationship between the location type and number of bicycle theft for each year. It clearly shows that bicycles were mostly stolen when parked outside, and this aligns with common sense that outdoor parking is more vulnerable to theft. However, it is surprising that the second most common location for bicycle theft is apartment instead of house. Residents' transportation habits may be a possible reason that since apartments are usually located in the town center and close to public transportation stations, bicycles is a good choice for short trips. For residents in house, they may rather choose vehicles that are faster such as cars to commute.

@fig-location_count_by_year also illustrates that the number of offence happened outside decreased while thefts happened indoor especially in residence increased in recent years. Factors contributing to the phenomenon could be further explored combining with other data.

It is noticeable that there are many reports with location type `Other`. Since it is not specified and the amount of `Other` type location cannot be neglected, how we treat the data may influence the analysis, and there may exists possible reporting bias.

```{r}
#| label: fig-location_count_by_year
#| fig-cap: Number of bicycle theft in different type of location in Toronto for each year from 2014 to 2022.
#| echo: false
#| warning: false
#| message: false

data |> ggplot(aes(x=location, fill=occ_year)) +
  geom_bar() +
  labs(x = "Location Type", y = "Number of occurrence", fill = "Occurrence Year") +
  scale_x_discrete() +
  theme_classic() +
  theme(axis.text.x=element_text(size=6))

@@

Other than external pattern, we are also interested how internal variables associated with bicycle thefts. @fig-diff_count_by_year shows the monthly difference in days between occurrence date and report date for each year from 2014 to 2022. It illustrates that people tend to report the offence within the same day of occurrence since there are more scatters near 0. Combined with @fig-yearly_count_by_month, there are much fewer thefts happened in December, however, the interval between occurrence and report date is much shorter in December since there are much more dots representing December near 0.

```{r}
#| label: fig-diff_count_by_year
#| fig-cap: Scatter plot of monthly difference in days between occurrence date and report date for bicycle thefts in Toronto for each year from 2014 to 2022.
#| echo: false
#| warning: false
#| message: false

data |> 
  ggplot(aes(x=occ_year, y=difference,
             color=occ_month)) +
  geom_point(alpha = 0.5) +
  labs(x = "Occurrence Year", y = "Difference in days between occurrence date and report date", fill = "Occurrence Month") +
  theme_classic()
```

@@

@tbl-rate_by_month shows that the recovery rate in winter is lower than that in summer in general. Consider potential seasonal pattern for bicycle thefts in a year, this may also be related to the weather in Toronto that severe and freezing weather in winter constrains people's activity in both positive and negative way. Even recovery rate is higher in summer, it is still not a high value that the largest recovery rate for each year is 17.58% and for each month is 15.70%.

@@

data |> group_by(occ_year) |> 
  summarise(rate = percent(sum(bicycle_status == "RECOVERED")/nrow(data)*100)) |> 
  arrange(rate) |> 
  kable(col.names = c("Year", "Recovery Rate"))

@@

data |> group_by(occ_month) |> 
  summarise(rate = percent(sum(bicycle_status == "RECOVERED")/nrow(data)*100)) |> 
  arrange(rate) |> 
  kable(col.names = c("Month", "Recovery Rate"))

@@

>>>>

The remainder of this paper is structured as follows: @sec-data discusses the data with @sec-data_introduction includes basic information about data collection and cleaning results, and @sec-data_visualization includes graphs and tables representing relationships between variables and some discussions; @sec-discussion discusses findings from data visualization and implications.

@@

```{r}
#| label: fig-yearly_count
#| fig-cap: Number of occurrence of bicycle theft in Toronto for each year from 2014 to 2022
#| echo: false
#| warning: false
#| message: false

data |> 
  mutate(occ_month = factor(occ_month, levels = month.name)) |> 
  ggplot(aes(x=as.factor(occ_year))) +
  geom_bar(color="darkgreen", fill="pink2") +
  labs(x = "Occurrence Year", y = "Number of occurrence", fill = "Occurrence Month") +
  theme_minimal()
```

@fig-yearly_count shows the total number of bicycle thefts in Toronto from 2014 to 2022. We could find that the total number of offence in a year increased before 2018 and began decrease since 2018. The change is relatively not significant within 1000, and the number of offence in 2022 is similar to that in 2014.

@fig-yearly_count_by_month illustrates potential seasonal pattern that there are much fewer bicycle thefts in winter, eg. December, January, and February. The offence increases a lot in other seasons especially in summer, eg. June, July, and August. It also shows that the season in which bicycles were stolen differs from year to year, but overall centers in the summer seasons and the seasonal difference is significant.

@@

data |> 
  mutate(occ_month = factor(occ_month, levels = month.name)) |> 
  group_by(occ_year, occ_month) |>
  summarise(n=n()) |>
  ggplot(aes(x = occ_month, y = n, group = occ_year, color =  as.factor(occ_year))) +
  geom_line(linewidth = 0.5) +
  theme_minimal() +
  labs(x = "Occurrence month", y = "Number of Occurrence", color = "Occurence year") +
  theme(axis.text.x=element_text(size=6.5)) 
```

Other than seasonal pattern, @tbl-location_count demonstrates the relationship between the location type and number of bicycle theft for each year. It clearly shows that bicycles were mostly stolen when parked outside, and this aligns with common sense that outdoor parking is more vulnerable to theft. However, it is surprising that the second most common location for bicycle theft is apartment instead of house.

@fig-location_count_by_year also illustrates that the number of offence happened outside decreased while thefts happened indoor especially in residence increased in recent years. Factors contributing to the phenomenon could be further explored combining with other data. It is noticeable that there are many reports with location type `Other`. Since it is not specified and the amount of `Other` type location cannot be neglected, how we treat the data may influence the analysis, and there may exists possible reporting bias.

```{r}
#| label: tbl-location_count
#| tbl-cap: Number of bicycle theft in different type of location in Toronto for each year from 2014 to 2022.
#| echo: false
#| warning: false
#| message: false

data |> group_by(location) |>
  summarise(n=n()) |>
  kable(col.names = c("Location type","Total number of occurrence"))
```
```{r}
#| label: fig-location_count_by_year
#| fig-cap: Yearly number of bicycle theft in different type of location in Toronto for each year from 2014 to 2022.
#| echo: false
#| warning: false
#| message: false

data |>
  group_by(location, occ_year) |>
  summarise(n=n()) |>
  ggplot(aes(x=location, y=n, group=occ_year, color=as.factor(occ_year))) +
  geom_line() +
  labs(x = "Location Type", y = "Number of occurrence", color = "Occurrence Year") +
  scale_x_discrete() +
  theme_minimal() +
  theme(axis.text.x=element_text(size=6))

@@

Other than external pattern, we are also interested how internal variables associated with bicycle thefts. @fig-diff_count_by_year shows the monthly difference in days between occurrence date and report date for each year from 2014 to 2022. It illustrates that people tend to report the offence within the same day of occurrence regardless of years. Very few people report bicycle theft after one month. It also consistent with @fig-yearly_count_by_month that more bicycle theft happen in summer instead of winter. However, the difference in report date and occurrence data does not show a seasonal pattern, namely regardless of seasons, people tend to report within the same day.

```{r}
#| label: fig-diff_count_by_year
#| fig-cap: Monthly difference in days between occurrence date and report date for bicycle thefts in Toronto for each year from 2014 to 2022.
#| echo: false
#| warning: false
#| message: false

data |>
  filter(difference <= 100) |>
  ggplot(aes(x=difference, color=factor(occ_month, levels = month.name))) +
  geom_freqpoly() +
  facet_wrap(~occ_year) +
  theme_minimal() +
  labs(x = "Difference in days", y = "Count", color = "Occurrence month")
```

@@

@tbl-rate_by_month shows that the recovery rate in winter is lower than that in summer in general. Consider potential seasonal pattern for bicycle thefts in a year, this may also be related to the weather in Toronto that severe and freezing weather in winter constrains people's activity in both positive and negative way. Even recovery rate is higher in summer, it is still not a high value that the largest recovery rate for each year is 17.6% and for each month is 15.7%.

@@

data |> group_by(occ_year) |> 
  summarise(rate = percent(round(sum(bicycle_status == "RECOVERED")/nrow(data)*100,3))) |> 
  arrange(rate) |> 
  kable(col.names = c("Year", "Recovery Rate"))

@@

data |> group_by(occ_month) |> 
  summarise(rate = (percent(round(sum(bicycle_status == "RECOVERED")/nrow(data)*100,3)))) |> 
  arrange(rate) |> 
  kable(col.names = c("Month", "Recovery Rate"))

@@

# Discussion {#sec-discussion}

## Bicycle theft shows an overall decreasing trend in recent years

## Bicycle theft shows a seasonal pattern

Possible reasons include improvement of security equipment of the bike, the market behind bicycle theft influenced by pandemic, and people's usage habit of bike. These implications need more data to analyze and could be our further interest.

This may be related to the weather in Toronto that many people will not ride bikes in winter due to extremely freezing weather and frequent snowfall.

## Bicycle theft shows a geographical pattern

Residents' transportation habits may be a possible reason that since apartments are usually located in the town center and close to public transportation stations, bicycles is a good choice for short trips. For residents in house, they may rather choose vehicles that are faster such as cars to commute.
